---
title: "Clustering"
author: "Sina Sanei"
date: "8/29/2019"
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing Data
The data set includes growth rate computed as height/age , for a total of 550 observations. 
There are 157 different variation of red oaks (reffered as populations) and 13 test sites which conducted the experiments. 


```{r cars , warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
data <- read_csv("df.csv")
data <- as.matrix(data[,c(2,3)])
```

## Assessing Clustering Tendency
Before applying any clustering method on your data, it’s important to evaluate whether the data sets contains meaningful clusters (i.e.: non-random structures) or not.

```{r  warning=FALSE, message=FALSE}
library(factoextra)
library(clustertend)
library(ggplot2)
res <- get_clust_tendency(data[sample(nrow(data),1000),], n = 1000-1, graph = FALSE)
res$hopkins_stat
```
It can be seen that our set is highly clusterable (the H value = 0.1710 which is well below the threshold= 0.5).  

visualizing the distance metrics : 


```{r warning=FALSE, message=FALSE}
dist.euc <- get_dist(data, stand = FALSE, method = "euclidean")

#fviz_dist(dist.euc , show_labels =FALSE )
  #, gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07"))

dist.man <- get_dist(data, stand = FALSE, method = "manhattan")

#fviz_dist(dist.man) 
  # gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07"))

#dist.cor <- get_dist(data.clust,  stand = FALSE, method = "pearson")
#fviz_dist(dist.cor, show_labels =FALSE,
         # gradient = list(low = "#380474", mid = "white", high = "#FC4E07"))
```


## K-means Clustering

The k-means clustering requires the users to specify the number of clusters to be generated.
for the euclidean distance metric:

```{r warning=FALSE, message=FALSE}
fviz_nbclust(data.clust,diss = dist.euc, kmeans, method = "wss" , k.max = 15) +
  geom_vline(xintercept = 5, linetype = 2)

fviz_nbclust(data.clust,diss = dist.euc, kmeans, method = "silhouette" , k.max = 15) +
  geom_vline(xintercept = 6, linetype = 2)

fviz_nbclust(data.clust,diss = dist.euc, kmeans, method = "gap_stat" , k.max = 45) +
  geom_vline(xintercept = 6, linetype = 2)
```

for the manhattan distance metric:

```{r warning=FALSE, message=FALSE}

dist.man <- get_dist(data.clust, stand = FALSE , method = "manhattan")

fviz_nbclust(data.clust, kmeans, method = "wss",  diss = dist.man , k.max = 15)  +
  geom_vline(xintercept = 5 , linetype = 2)

fviz_nbclust(data.clust, kmeans, method = "silhouette",  diss = dist.man , k.max = 10)  
```

for the pearson distance metric:

```{r warning=FALSE, message=FALSE}
fviz_nbclust(data.clust, kmeans, method = "wss",  diss = dist.cor , k.max = 25)  +
  geom_vline(xintercept =c(2, 5), linetype = 2)

fviz_nbclust(data.clust, kmeans, method = "silhouette",  diss = dist.cor , k.max = 25)  +
  geom_vline(xintercept =c(2, 5), linetype = 2)
```

Compute k-means with k = 5

```{r warning=FALSE, message=FALSE}
set.seed(17)
km.euc <- kmeans(data, 6 , nstart = 25)
print(km.euc)

aggregate(data.clust, by=list(cluster=km.euc$cluster), mean)
```
  
Visualizing k-means with k=6:  

```{r warning=FALSE, message=FALSE}
fviz_cluster(km.euc, data = data.clust,
ellipse.type = "euclid", # Concentration ellipse 
star.plot = TRUE, # Add segments from centroids to items 
ggtheme = theme_minimal()
)

fviz_cluster(list(data = data.clust, cluster = km.euc$cluster),
             ellipse.type = "convex", geom = "point", stand = FALSE,
             palette = "jco", ggtheme = theme_classic())
```


Notice that kmeans is sensitive to outliers , we can try to use PAM (Partitioning Around Medoids, Kaufman & Rousseeuw, 1990) algorithm, which is less sensitive to outliers.

## K-Medoids Clustering  

In k-medoids clustering, each cluster is represented by one of the data point in the cluster. These points are named cluster medoids.   
Estimating the optimal number of clusters:  

```{r warning=FALSE, message=FALSE}
library(cluster)
fviz_nbclust(data.clust, pam, method = "silhouette")+geom_vline(xintercept =6 , linetype = 2) + theme_classic()

fviz_nbclust(data.clust, pam, method = "wss")+geom_vline(xintercept =6 , linetype = 2) + theme_classic()
```
  
computing PAM algorithm with k = 2, euclidean dist :


```{r warning=FALSE, message=FALSE}
pam.euc <- pam(data.clust, 5, metric = "euclidean") 
print(pam.euc)

fviz_cluster(pam.euc,
ellipse.type = "convex", # Concentration ellipse 
ggtheme = theme_classic()
)
```

using Manhattan distance :

```{r warning=FALSE, message=FALSE}
pam.man <- pam(data.clust, 6, metric = "manhattan") 
print(pam.man)

fviz_cluster(pam.man,
ellipse.type = "convex", # Concentration ellipse 
ggtheme = theme_classic()
)
```  


## DETERMINING THE OPTIMAL NUMBER OF CLUSTERS  
There are three different methods to determine the optimal number of clusters: 
ELbow method, Average Silhouette method and Gap statistic method.

```{r warning=FALSE, message=FALSE}
fviz_nbclust(data.clust, kmeans, method = "wss") +
   geom_vline(xintercept = 5, linetype = 2) +
   labs(subtitle = "Elbow method")
```  

```{r warning=FALSE, message=FALSE}
fviz_nbclust(data.clust, kmeans, method = "silhouette") +
   geom_vline(xintercept = 6, linetype = 2) +
   labs(subtitle = "Silhouette method")
```  

```{r warning=FALSE, message=FALSE}
set.seed(17)
fviz_nbclust(data.clust, pam , method = "gap_stat", nboot = 50)+
labs(subtitle = "Gap statistic method")
``` 

```{r warning=FALSE, message=FALSE}
library(NbClust)
num <- NbClust(data = data.clust, distance = "euclidean",
              min.nc = 2, max.nc = 15, method = "kmeans")
fviz_nbclust(num)
``` 
```{r warning=FALSE, message=FALSE}
num <- NbClust(data = data.clust, distance = "maximum",
              min.nc = 2, max.nc = 15, method = "kmeans")
fviz_nbclust(num)
```
```{r warning=FALSE, message=FALSE}
num <- NbClust(data = data.clust, distance = "manhattan",
              min.nc = 2, max.nc = 15, method = "kmeans")
fviz_nbclust(num)
```
```{r warning=FALSE, message=FALSE}
num <- NbClust(data = data.clust, distance = "manhattan",
              min.nc = 2, max.nc = 15, method = "median")
fviz_nbclust(num)
```


```{r warning=FALSE, message=FALSE}
num <- NbClust(data = data.clust, distance = "euclidean",
              min.nc = 2, max.nc = 15, method = "median")
fviz_nbclust(num)
```

```{r warning=FALSE, message=FALSE}
num <- NbClust(data = data.clust, distance = "euclidean",
              min.nc = 2, max.nc = 15, method = "centroid")
fviz_nbclust(num)
```

```{r warning=FALSE, message=FALSE}
num <- NbClust(data = data.clust, distance = "manhattan",
              min.nc = 2, max.nc = 15, method = "centroid")
fviz_nbclust(num)
```

```{r warning=FALSE, message=FALSE}
num <- NbClust(data = data.clust, diss=dist.cor ,distance = NULL,
              min.nc = 2, max.nc = 15 , method = "centroid")
fviz_nbclust(num)
```

```{r warning=FALSE, message=FALSE}
num <- NbClust(data = data.clust, diss=dist.cor ,distance = NULL,
              min.nc = 2, max.nc = 15 , method = "kmeans")
fviz_nbclust(num)
```

```{r warning=FALSE, message=FALSE}
num <- NbClust(data = data.clust, diss=dist.cor ,distance = NULL,
              min.nc = 2, max.nc = 15 , method = "ward.D2")
fviz_nbclust(num)
```  



```{r warning=FALSE, message=FALSE}
library(ggmap)
library(ggConvexHull)
loc.data <- data[,1:71] 
loc.data <- loc.data %>% filter(popID != "752")
loc <- c(mean(loc.data$long.ss), mean(loc.data$lat.ss))
my_map <- get_map(c(left = -99, bottom = 32, right = -60, top = 55), zoom =5 , scale =1, maptype = 'terrain-lines')
loc.data$clust <-pam.euc$clustering
ggmap(my_map) +
  geom_point(aes(long.ss, lat.ss, color=factor(clust)),shape =6,size =2,alpha=0.6, data = loc.data) +
  geom_convexhull(data = loc.data,alpha=.5,aes(long.ss, lat.ss, fill=factor(clust), color=factor(clust)) )
``` 